Inferring stellar rotation periods using Gaussian processes

### 2017-01-26 documents/ms.tex
updating plots and adding words about kois

\documentclass[useAMS, usenatbib, preprint, 12pt]{aastex}
\usepackage{cite, natbib}
\usepackage{float}
\usepackage{epsfig}
\usepackage{cases}
\usepackage[section]{placeins}
\usepackage{graphicx, subfigure}
\usepackage{color}
\usepackage{bm}
\usepackage{enumerate}

\newcommand{\columbia}{1}
\newcommand{\oxford}{3j}
\newcommand{\uw}{4}
\newcommand{\princeton}{2}
\newcommand{\naigrain}{333}
\newcommand{\nmcquillan}{100}
\newcommand{\nkois}{1114}
\newcommand{\nkoimcq}{271}
\newcommand{\nkoigood}{222}
\newcommand{\nkoibad}{49}
\newcommand{\kepexample}{1430163}
\newcommand{\kepexampleperiod}{4}
\newcommand{\aigrainexampleperiod}{20.8}
\newcommand{\Kepler}{{\it Kepler}}
\newcommand{\kepler}{\Kepler}
\newcommand{\corot}{{\it CoRoT}}
\newcommand{\Ktwo}{{\it K2}}
\newcommand{\ktwo}{\Ktwo}
\newcommand{\TESS}{{\it TESS}}
\newcommand{\LSST}{{\it LSST}}
\newcommand{\Wfirst}{{\it Wfirst}}
\newcommand{\SDSS}{{\it SDSS}}
\newcommand{\PLATO}{{\it PLATO}}
\newcommand{\Teff}{$T_{\mathrm{eff}}$}
\newcommand{\teff}{$T_{\mathrm{eff}}$}
\newcommand{\FeH}{[Fe/H]}
\newcommand{\feh}{[Fe/H]}
\newcommand{\ie}{{\it i.e.}}
\newcommand{\eg}{{\it e.g.}}
\newcommand{\logg}{log \emph{g}}
\newcommand{\dnu}{$\Delta \nu$}
\newcommand{\numax}{$\nu_{\mathrm{max}}$}
\newcommand{\acfRMS}{9.27}
\newcommand{\pgramRMS}{1.25}
\newcommand{\mcmcRMS}{0.46}

@@

\author{%
   Ruth Angus\altaffilmark{\columbia},
   Timothy Morton\altaffilmark{\princeton}
   Suzanne Aigrain\altaffilmark{\oxford},
   \& Daniel Foreman-Mackey\altaffilmark{\uw}
}

@@

\begin{figure*}
\begin{center}
\includegraphics[width=6in, clip=true]{figures/compare_pgram.pdf}
\caption[LS periodogram results.]
{The `true' rotation periods used to generate \naigrain\ simulated light
curves vs the rotation periods measured using a LS periodogram technique.}
\label{fig:pgram_compare}
\end{center}
\end{figure*}

@@

In order to manage the great flexibilty of the GP model and focus on 
reliably retrieving the period parameter, we also impose priors
on the non-period GP parameters that we
learned during the development of this method.
Beginning with very broad priors (uniform in log space between
-20 and 20) on all the parameters, we first explored what the
typical values of the non-period parameters tended to be when the
injected periods were successfully recovered.  We also experimented with
constraining the allowed ranges of the parameters, after discovering
that some regions of parameter space (such as large values of $A$ and $\Gamma$
and small values of $l$) tended to allow fits that ignored the desired
periodicity.  The priors and bounds we used for this work are listed in
table \ref{tab:priors}.  We note that we do not have any specific
quantitative justification for the use of these priors,
except that they produce good results for this particular test set of
light curves.  We thus caution that other data sets might have different
noise properties, potentially requiring modifications to these priors
on the non-period parameters.  For period, we tried two different methods:
an uninformed (log-flat) prior between 0.5\,d and 100\,d, and an ACF-informed
prior (details in Appendix.)
% We note that our default prior for period is
% log-uniform between the bounds, but that we were also able to
% further improve our results by using an ACF-inspired prior.

To sample the posterior in a way that is sensitive to potential
multimodality, we use the \texttt{emcee3} MCMC sampler.
We initialize 500 walkers with random samples from the prior
 and use a mix of different proposal distributions:
specifically, a 2:2:1 mix of KDE, differential evolution, and DESnooker moves.
[DFM: more words/citations here?]
We run fifty steps of the sampler at a time, checking for convergence after
each iteration, up to a maximum of fifty iterations.  We declare convergence if
the total effective chain length is at least 8$\times$ the maximum autocorrelation
time.  When convergence is achieved, we drop the first two autocorrelation 
lengths in the chain as a burn-in, and randomly choose 5000 samples as 
representative of the posterior.  This fitting process takes several hours 
for a typical simulated light curve, though in some cases it can take 12
hours or longer to converge.

\begin{table*}
\begin{center}
\caption{Priors and bounds on the natural logarithms of the GP model parameters.}
\begin{tabular}{lcc}
Parameter & Prior & Bounds\\
    \hline
    $\ln A$ & $\mathcal N(-13, 5.7)$ & (-20, 0) \\
    $\ln l$ & $\mathcal N(7.2, 1.2)$ & (2, 20) \\
    $\ln \Gamma$ & $\mathcal N(-2.3, 1.4)$ & (-10, 3) \\
    $\ln \sigma$ & $\mathcal N(-17, 5)$ & (-20, 0) \\
    $\ln P $ & Uniform / ACF-based & ($\ln 0.5, \ln 100$) \\ 
\end{tabular}
\end{center}
\end{table*}
\label{tab:priors}

@@

Figures \ref{fig:compare_mcmc_noprior} and \ref{fig:compare_mcmc_acfprior} 
summarize the results of this MCMC fitting 
procedure compared to the injected ``true'' stellar rotation periods, for the 
uninformed and ACF-informed priors on period, respectively.
The marginal posterior distributions of the QP kernel hyperparameters for the
example simulated light curve with in figure \ref{fig:demo_lc}, are shown in
figure \ref{fig:gp_posteriors}.

@@

In order to test this rotation period inference method on real data,
we applied it to a set of \nkois\ \Kepler\ Object of Interest (KOI)
host stars, \nkoimcq\ of which had rotation periods previously measured by 
the ACF method by \citet{Mcquillan2013}.  We use the pipeline-corrected flux 
(\texttt{pdcsap_flux} column in the \Kepler\ light curve table), median-normalized
and unit-subtracted.  As with the simulated light curves, we randomly subsample each 
light curve by a factor of 30 and split it into segments of about 300 points
for the purposes of evaluating the likelihood.  We also follow the same MCMC
fitting procedure as with the simulated data.

Figure \ref{fig:kois} compares the periods inferred with the GP method to the
ACF-based periods from \citet{Mcquillan2013} for the \nkoimcq\ overlapping KOIs.
This comparison shows generally very good agreement, 
with the exception of a cluster of stars in the lower right of the figure
where the GP method does not recover the ACF period.  
However, rather than being primarily evidence of weakness of the GP method,
this cluster of nominal failures actually reveals an important advantage
for this method---the possibility of quantitatively motivated quality assessment.

@@

Our results show good agreement with the \citet{Mcquillan2014} measurements,
as shown in \ref{fig:mcquillan}.

\begin{figure}
\begin{center}
\includegraphics[width=6in, clip=true]{figures/comparison_koi.pdf}
\caption[Comparison with McQuillan results.]
{A comparison of our GP rotation period measurements to those of
\citet{Mcquillan2014}. The results show good agreement. In several cases we
    measure a rotation period that is half the value reported by
    \citet{Mcquillan2014}.}
\label{fig:mcquillan}
\end{center}
\end{figure}

@@

We used the \naigrain\ simulated \kepler-like light curves for solid-body
rotation from \citet{Aigrain2015} and attempted to recover the rotation
periods used to generate them.
Three methods were compared: a Lomb-Scargle periodogram method, an
autocorrelation function method and a new Gaussian process method.
The GP method produced the most precise and accurate rotation periods of the
three techniques.
This is because a GP is a semi-parametric model that is well suited to signals
that are non-sinusoidal and quasi-periodic.
Not only this, unlike the other two methods, the GP method provides posterior
PDF samples of rotation periods, allowing characterisation of uncertainties.
Note that the posterior PDFs are often non-Gaussian, being bi- or multimodal.
In these cases it is therefore misleading to report a 1$\sigma$ or 16th and
84th percentiles as uncertainties.
Rather, it is better to incorporate posterior PDF samples into analyses which
require uncertainty propagation.

Because rotation periods are inferred probabilistically with the GP method,
this allows them to be easily incorporated into follow-on studies; inferring
stellar ages or characterising star-planet interactions for example.
This the GP method also eliminates the need to define a heuristic detection
criterion such as are often used in the literature: a light curve that does
not contain a rotation signal will return the prior PDF.

The main drawback of the GP method is the computation time.

>>>>

\documentclass[useAMS, usenatbib, preprint, 12pt]{aastex}
\usepackage{cite, natbib}
\usepackage{float}
\usepackage{epsfig}
\usepackage{cases}
\usepackage[section]{placeins}
\usepackage{graphicx, subfigure}
\usepackage{color}
\usepackage{bm}

\newcommand{\columbia}{1}
\newcommand{\oxford}{3}
\newcommand{\uw}{4}
\newcommand{\princeton}{2}
\newcommand{\naigrain}{333}
\newcommand{\nmcquillan}{100}
\newcommand{\kepexample}{1430163}
\newcommand{\kepexampleperiod}{4}
\newcommand{\aigrainexampleperiod}{20.8}
\newcommand{\Kepler}{{\it Kepler}}
\newcommand{\kepler}{\Kepler}
\newcommand{\corot}{{\it CoRoT}}
\newcommand{\Ktwo}{{\it K2}}
\newcommand{\ktwo}{\Ktwo}
\newcommand{\TESS}{{\it TESS}}
\newcommand{\LSST}{{\it LSST}}
\newcommand{\Wfirst}{{\it Wfirst}}
\newcommand{\SDSS}{{\it SDSS}}
\newcommand{\PLATO}{{\it PLATO}}
\newcommand{\Teff}{$T_{\mathrm{eff}}$}
\newcommand{\teff}{$T_{\mathrm{eff}}$}
\newcommand{\FeH}{[Fe/H]}
\newcommand{\feh}{[Fe/H]}
\newcommand{\ie}{{\it i.e.}}
\newcommand{\eg}{{\it e.g.}}
\newcommand{\logg}{log \emph{g}}
\newcommand{\dnu}{$\Delta \nu$}
\newcommand{\numax}{$\nu_{\mathrm{max}}$}
\newcommand{\acfRMS}{9.27}
\newcommand{\pgramRMS}{1.25}
\newcommand{\mcmcRMS}{0.46}

@@

\author{%
   Ruth Angus\altaffilmark{\columbia},
   Timothy Morton\altaffilmark{\princeton}
   Suzanne Aigrain\altaffilmark{\oxford},
   Daniel Foreman-Mackey\altaffilmark{\uw}
   Suzanne Aigrain\altaffilmark{\oxford},
   \& Vinesh Rajpaul\altaffilmark{\oxford}
}

@@

\begin{figure*}
\begin{center}
\includegraphics[width=6in, clip=true]{figures/compare_pgram.pdf}
\caption[LS periodogram results.]
{The `true' rotation periods used to generate \naigrain\ simulated light
curves vs the rotation periods measured using a LS periodogram technique.
In many cases a large peak at a long period was present in the periodogram,
    producing a significant over-estimate of the period.
    }
\label{fig:pgram_compare}
\end{center}
\end{figure*}

@@

In order to manage the extreme flexibilty of the GP model and focus on
reliably retrieving the period parameter, we also impose priors
on the non-period GP parameters that we
learned during the development of this method.
Beginning with very broad priors (uniform in log space between
-20 and 20) on all the parameters, we first explored what the
typical values of the non-period parameters tended to be when the
injected periods were successfully recovered.  We also experimented with
constraining the allowed ranges of the parameters, after discovering
that some regions of parameter space (such as large values of $A$ and $\Gamma$
and small values of $l$) tended to allow fits that ignored the desired
periodicity.  The priors we used for this work are listed in
table \ref{tab:priors}.  We note that we do not have any particular
quantitative justification for the use of these priors,
except that they produce good results for this particular test set of
light curves.  We thus caution that other data sets might have different
noise properties, potentially requiring modifications to these priors.
% We note that our default prior for period is
% log-uniform between the bounds, but that we were also able to
% further improve our results by using an ACF-inspired prior.

To sample the posterior in a way that is sensitive to potential
multimodality, we use the \texttt{emcee3} MCMC sampler.
We initialize 500 walkers with random samples from the prior
 and use a mix of different proposal distributions:
specifically, a 2:2:1 mix of KDE, differential evolution, and DESnooker moves.
[DFM: more words/citations here?]
We run fifty steps of the sampler at a time, checking for convergence after
each iteration, up to a maximum of fifty iterations.  We declare convergence if
the total effective chain length is at least 8$\times$ the maximum autocorrelation
time.  When convergence is achieved, we drop the first two autocorrelation
lengths in the chain as a burn-in, and randomly choose 5000 samples as
representative of the posterior.  This fitting process takes several hours
for a typical simulated light curve.

\begin{table*}
\begin{center}
\caption{Priors and bounds on the natural logarithms of the GP model parameters.}
\begin{tabular}{lcc}
Parameter & Prior & Bounds\\
    \hline
    $\ln A$ & $\mathcal N(-13, 5.7)$ & (-20, 0) \\
    $\ln l$ & $\mathcal N(7.2, 1.2)$ & (2, 20) \\
    $\ln \Gamma$ & $\mathcal N(-2.3, 1.4)$ & (-10, 3) \\
    $\ln \sigma$ & $\mathcal N(-17, 5)$ & (-20, 0) \\
    $\ln P $ & Uniform & ($\ln 0.5, \ln 100$) \\
\end{tabular}
\end{center}
\end{table*}
\label{tab:priors}

@@

Figure \ref{fig:compare_mcmc} summarizes the results of this MCMC fitting procedure
compared to the injected ``true'' stellar rotation periods.
The marginal posterior distributions of the QP kernel hyperparameters for the
example simulated light curve with in figure \ref{fig:demo_lc}, are shown in
figure \ref{fig:gp_posteriors}.

@@

@@

Our results show good agreement with the \citet{Mcquillan2014} measurements,
as shown in \ref{fig:mcquillan}.
The uncertainties published in the \citet{Mcquillan2014} catalogue were
calculated from the scatter in the positions of peaks in the ACF that are
separated by integer multiples of the period.
We show a comparison of our results to those of the \kepler\ Objects of
Interest (KOIS) \citet{Mcquillan2014} in order to demonstrate that our
technique works on real \kepler\ light curves, not just simulated ones.
These are perhaps not the ideal test subjects because many KOI light curves
have deep stellar occulatations and planetary transits which can look
like non-Gaussian noise.
Since Gaussian processes require that the data are Gaussian, this could have
detremental effects on the results.
Despite this however, our results are, in most cases, similar to those of
\citet{Mcquillan2014} indicating that our method still works well even in the
presence of transits and eclipses.
In many cases where there is a large discrepancy between our results and the
\citet{Mcquillan2014} results, either the variability is low amplitude or
frequent, deep eclipses/transits are present in the data.
In most other cases we expect our results to be the more accurate ones and the
discrepancy may be due to quasi-periodicity in the light curve.

\begin{figure}
\begin{center}
\includegraphics[width=6in, clip=true]{figures/comparison_koi.pdf}
\caption[Comparison with McQuillan results.]
{A comparison of our GP rotation period measurements to those of
\citet{Mcquillan2014}.
The data points are coloured by the range of variability measured by
    \citet{Mcquillan2014}, defined as the interval between the 5th and 95th
    percentiles of normalized flux per period bin in millimagnitudes.
    The results show good agreement.}
\label{fig:mcquillan}
\end{center}
\end{figure}

@@

We used the \naigrain\ simulated \kepler-like light curves for solid-body
rotation from \citet{Aigrain2015} and attempted to recover the rotation
periods used to generate them.
Three methods were compared: a Lomb-Scargle periodogram method, an
autocorrelation function method and a new Gaussian process method.
We demonstrate that the GP method produced the most precise and accurate
rotation periods of the three techniques.
In addition, we compared the rotation periods of \kepler\ objects of
interest measured using the GP method to those measured previously by
\citet{Mcquillan2014}.
The good agreement between the two sets of results demonstrates that the GP
method works well on real \kepler\ data and is, in most cases robust to the
presents of transits and eclipses in the light curves.

Although rotation period uncertainties calculated using the GP method are
better estimates than those calculated using the ACF method, they are still
imperfect estimates of the intrinsic uncertainties.
This is because the posterior PDF of the rotation period parameter is, in many
cases, multi-modal.
If it were possible to run an MCMC sampler for an infinite number of steps, we
would the entire posterior PDF would eventually be sampled, producing an
accurate uncertainty measurement.
Practically speaking however, these imperfect estimates may be the best
currently achievable.

A GP is a semi-parametric model that is well suited to signals that are
non-sinusoidal and quasi-periodic.
Not only this, unlike the other two methods, the GP method provides posterior
PDF samples of rotation periods, allowing characterisation of uncertainties.
Note that the posterior PDFs are often non-Gaussian, being bi- or multimodal.
In these cases it is therefore misleading to report median values or 16th and
84th percentiles as uncertainties.
Rather, it is better to incorporate posterior PDF samples into analyses which
require uncertainty propagation.
Because rotation periods are inferred probabilistically with the GP method,
this allows them to be easily incorporated into follow-on studies; inferring
stellar ages or characterising star-planet interactions for example.

The main drawback of the GP method is the computational cost.
